Simple repression draft

### 2020-05-30 doc/appendix_03_bayesian_inference.tex
MRM cleaning inference section in SI. Still needs input from MM

The structure of the data in~\cite{Jones2014} for regulated promoters tuned
these two parameters independently. In their work the production of the LacI
repressor was under the control of an inducible promoter regulated by the TetR
repressor as schematized in Figre~\ref{figS:aTc_circuit}. When TetR binds to the
small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
unable to bind to the DNA. This translates into an increase in gene expression
level. In other words, the higher the concentration of aTc added to the media,
the less TetR repressors that can control the expression of the \textit{lacI}
gene, so the higher the concentration of LacI repressors in the cell.

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes
the ``right'' model of a constitutive promoter is the bursty
picture, model five in Figure~\ref{fig:constit_cartoons}.
Therefore our starting point here is the analogous model with
repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
For a given repressor binding site and copy number,
this model has four rate parameters to be inferred:
the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
the initiation rate of bursts, $k_i$, and the mean burst size $b$
(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical
inference model, we would like to sketch the intuitive structure.
The dataset from~\cite{Jones2014} we consider consists of
single-cell mRNA FISH measurements of nine different conditions,
spanning several combinations of three unique binding sites and
four unique repressor copy numbers. We assume that the values of
$k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that
these values are the same across datasets with different
repressor binding sites and copy numbers. We assume that there is
one unbinding rate parameter for each repressor binding site,
which is shared across conditions, and likewise one binding rate
for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note the repressor copy numbers are not known directly, so we label their
association rates by the concentration of a small molecule inducer aTc
which indirectly controls the transcription of $\textit{lacI}$.
Also note that the authors of~\cite{Jones2014} report estimates
of LacI copy number per cell rather than direct measurements.
However, these estimates were made \textit{assuming} the validity
of the equilibrium models in Figure~\ref{fig1:means_cartoons},
and since \textit{testing} these models is our goal, we will make
no assumptions about the LacI copy number for given aTc concentrations.

Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

\mmnote{Probably SI most of this paragraph?}
This likelihood is rather inscrutable. We did not find any of the
known analytical approximations for ${_2F_1}$ terribly useful in
gaining intuition, so we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
with burst size $b$ and burst rate $k_i$ equal to their
constitutive \textit{lacUV5} values, except with $k_i$ multiplied
by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates
that the repressor rates had left an imprint on the distribution,
and perhaps intuitively, this intermediate regime occurs for values of
$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
Only when all three rates in the problem are comparable does the
mRNA distribution retain detectable information about them.

@@

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
\mmnote{Now that I've typed this out, this notation seems totally
pointless here, especially when the likelihood is not a standard
distribution. Maybe I should make a table or something to record
the prior values, and maybe even that should go to SI?}
We ran MCMC sampling on the full nine dimensional posterior
specified by this generative model. To attempt to visualize this
object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
several two-dimensional slices as contour plots, analogous to
Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
conditions from the dataset used to fit the model and gives a
sense of the uncertainty and correlations in the posterior.

>>>>

The structure of the data in~\cite{Jones2014} for regulated promoters tuned
these two parameters independently. In their work the production of the LacI
repressor was under the control of an inducible promoter regulated by the TetR
repressor as schematized in Figre~\ref{figS:aTc_circuit}. When TetR binds to the
small molecule anhydrotetracycline (aTc), it shifts to an inactive conformation
unable to bind to the DNA. This translates into an increase in gene expression
level. In other words, the higher the concentration of aTc added to the media,
the less TetR repressors that can control the expression of the \textit{lacI}
gene, so the higher the concentration of LacI repressors in the cell. So by 
tuning the amount of aTc in the media where the experimental strains were grown
they effectively tune $k^+$ in our simple theoretical model. On the other hand
to tune $k^-$ the authors swap three different binding sites for the LacI 
repressor, each with different repressor-DNA binding affinities previously 
characterized \cite{Garcia2011a}.

@@

What this means is that we have access to data with different combinations of
$k^-$ and $k^+$. We could naively try to fit the kinetic parameters individually
for each of the datasets, but there is no reason to believe that the binding
site identity for the LacI repressor somehow affects its expression level
controlled from a completely different location in the genome, nor vice versa.
In other words, what makes the most sense it to fit all datasets together to
obtain a single value for each of the association and dissociation rates. What
this means, as described in~\ref{section_04_bayesian_inference} of the main text
is that we have a seven dimensional parameter space with four possible
association rates $k^+$ given the four available aTc concentrations, and three
possible dissociation rates $k^-$ given the three different binding sites
available in the dataset.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note that since the repressor copy numbers are not known directly as explained
before, we label their association rates by the concentration of aTc. Bayes
theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts across the
various conditions. We assume that individual single-cell measurements are
independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding rates for the
$j$-th measured cell. Our likelihood function, previously derived in
Appendix~\ref{sec:gen_fcn_appdx}, is given by the rather complicated result in
Eq.~\ref{eq:p_m_bursty+rep_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

\mrm{This paragraph needs its own subsection in my opinion if it is to be 
included in the SI.}
This likelihood is rather inscrutable. We did not find any of the known
analytical approximations for ${_2F_1}$ terribly useful in gaining intuition, so
we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial with
burst size $b$ and burst rate $k_i$ equal to their constitutive \textit{lacUV5}
values, except with $k_i$ multiplied by the fold-change
$\left(1+k_R^+/k_R^-\right)^{-1}$. In other words, once again only the
\textit{ratio} $k_R^+/k_R^-$ was detectable. But for intermediate repression,
the distribution was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates that the repressor
rates had left an imprint on the distribution, and perhaps intuitively, this
intermediate regime occurs for values of $k_R^\pm$ comparable to the burst rate
$k_i$. Put another way, if the repressor rates are much faster or much slower
than $k_i$, then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$. Only when all three
rates in the problem are comparable does the mRNA distribution retain detectable
information about them.

@@

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}. We ran MCMC
sampling on the full nine dimensional posterior specified by this generative
model.

We found that fitting a single operator/aTc concentration at a time with a
single binding and unbinding rate did not yield a stable inference for most of
the possible operator/aTc combinations. In other words, a single dataset could
not independently resolve the binding and unbinding rates, only their ratio as
set by the mean fold-change in Figure~\ref{fig1:means_cartoons} in the main
text. Only by making the assumption of a single unique binding rate for each
repressor copy number and a single unique unbinding rate for each binding site,
as done in Figure~\ref{fig4:repressed_post_full}(A), was it possible to
independently resolve the rates and not merely their ratios.

We also note that we found it necessary to exclude the very weakly and very
strongly repressed datasets from Jones et.\ al.~\cite{Jones2014}. In both cases
there was, in a sense, not enough information in the distributions for our
inference algorithm to extract, and their inclusion simply caused problems for
the MCMC sampler without yielding any new insight. For the strongly repressed
data (Oid, 10~ng/mL aTc), with $>$ 95\% of cells with zero mRNA, there was quite
literally very little data from which to infer rates. And the weakly repressed
data, all with the repressor binding site O3, had an unbinding rate so fast that
the sampler essentially sampled from the prior; the likelihood had negligible
influence, meaning the data was not informing the sampler in any meaningful way,
so no inference was possible.
